Learning from Data Through Models

### 2023-07-11 paper/paper.Rmd
edit framework writing

A finite number of observable random variables, $X =(x_1, x_2, ..., x_N)$, determine a binary state, $y$. 
The observable variables are independently distributed and each follows a Normal distribution with corresponding mean 
$\mu_i$ and standard deviation $\sigma_i$ with $i \in \{1, 2, ..., N\}$.

The state is \emph{Red} (**R**) or \emph{Blue} (**B**) and is determined in the following way:
\begin{equation*}
y = \begin{cases}
\text{\textbf{B}} &\text{if $a_1x_1+...+a_Nx_N + K\geq0$}\\
\text{\textbf{R}} &\text{otherwise}
\end{cases}
\end{equation*}

@@

In each period, the decision-maker (DM) observes the realization of all observable variables and predicts the state. 
Their flow payoff in period $t$ is 1 if they predict the state correctly, and it is zero otherwise. 
Denote the prediction by $\hat{y}$.

In order to make a prediction, the DM uses a map $f: M_t\to \{\textbf{B}, \textbf{R}\}$  
where $M_t \subseteq \{x_1, x_2, ..., x_N\}$ is the set of variables that they choose to consider in period $t$. 
Our main object of study is the set of variables that the DM decides to consider and not the properties of the 
mapping from the variables to the state. We refer to the set of variables that are taken into consideration, $M_t$, 
as the DM's *model of the world*, and we abstract away from how the variables are mapped into a prediction of 
the state.

A model of the world is said to be *simple* if it does not include all the relevant variables, 
*i.e.* if $M\subset \mathcal{R}$. And it is complex if it includes all relevant variables, $M=\mathcal{R}$. Thus, whenever 
agents are using simple models, they are ignoring some of the relevant variables even if the information is available to them.
In contrast, whenever agents are using complex models, they are using all the relevant information that is available to them, 
so restricting the number of variables that is available to them will impair their ability to predict correctly. In the design
of the experiment we will exploit these differences in order to determine whether subjects are using simple or complex models.

**Prediction 1:** If subjects are using simple model with $K$ variables, then allowing them to disclose $K+1$ variables 
will not increase the frequency with which they make correct predictions.

**Prediction 1B:** If subjects are using complex models, then allowing them to disclose an additional variable will always 
increase the frequency with which they make correct predictions.

### Optimal Rules
The most accurate prediction rule for this setting will be one that predicts the state to be **B** whenever, 
given the observables, **B** is more likely. And it predicts **R** whenever **R** is more likely 
given the observables. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the DM has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already.

We refer to a prediction rule as an *Optimal Rule* if it is the rule that a fully rational Decision maker would 
converge to with perfect Bayesian learning. In this context, Bayesian learning corresponds to using Bayes rule to 
learn the coefficients $a = (a_1, a_2, ..., a_N)$ as well as the constant $K$. This means that in an $N$ dimensional 
space, the Decision Maker must learn $N+1$ parameters which corresponds to learning the halfspace in which the state 
is **B**. The optimal rule for the complex model coincides with the true data-generating process when learning 
is perfect.

Conditional on using a simple model, the optimal rule for that model is the rule that, when restricted to the 
variables being considered, does the best at predicting the state. In this case, the Decision Maker is learning only 
the parameters that pertain the variables in the model. Hence, for a model that considers $K$ variables, the DM will 
have to learn $K+1$ parameters. The following proposition illustrates exactly what the optimal prediction rules look 
like when the DM is able to learn.

##### Proposition.
_Fix any model that considers $k\leq N$ variables and relabel the variable so that $x_1, ..., x_K$ are the variables 
that the model considers and $x_{k+1}, ..., x_N$ are the ones that the model does not consider. Relabel the 
coefficients $a = (a_1, ..., a_N)$ accordingly as well._

##### proof.
_Let $M:=a_1x_1+...+a_kx_k$ and $M^c = a_{k+1}x_{k+1}+...+a_Nx_N+k$ and define the latent variable $y^L := M +  M^c$. 
The optimal prediction rule for a model that considers the variables in $M$ the state to be $\textbf{B}$ whenever 
$P[y^L|M]\geq \frac{1}{2}$ and predicts $\textbf{R}$ otherwise._

@@

With limited data, it is not guaranteed that the DM is able to perfectly learn the parameters required for the optimal rule. 
However, the notion of optimal rules is useful for determining the scenarios in which we can expect polarization to arise. 
The next section uses the notion of optimal rules to formalize the concept of polarization and to establish the 
main prediction of the theory.

### Polarization
To study polarization, we take the definition of @Haghtalab2021: two DMs are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one DM predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in @Haghtalab2021---polarization may persist even when DMs have 
access to infinite data because they might be using different models of the world in order to make their predictions. 
Nevertheless using different models does not necessarily imply that beliefs will always be polarized; it must also be 
the case that the realization of observables falls in a *disagreement zone*. By disagreement zone, we simply 
mean that if the two models that are being used for prediction are $M_1$ and $M_2$, the realization of the 
observables $x=(x_1, ..., x_N)$ must be such that $M_1$ makes a different prediction from $M_2$. The following 
2-dimensional example illustrates why the use of different models is necessary but not sufficient for polarization 
to arise.

##### Example
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2+1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As the figure makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. Theregions in which the rules make different predictions
are the disagreement zones._

For any pair of agents who observe the same realization of teh observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by their models. Notice that the 
polarizing observations---those that fall into a polarization zone---are different for each pair of models. Also notice that 
our definition of the polarization zone assumes that the DMs are using optimal rules. This is a very strong assumption 
and will most likely not be the case that subjects in the experiment learn perfectly. However, the notion of optimal rules
allows us to have a clean definition of polarization and to make predictions about when polarization will arise.

@@

Subjects for the experimet were recruited from the CESS lab subject pool at NYU. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person at 
the CESS lab. The experiment was coded using o-tree @otree and it consisted of two parts: Part 1 was meant to expose the 
subjects to the data generating process
and allow them to learn how to predict; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

Upon arrival subjects were randomly placed in pairs, however, they were not informed of this fact.
Part 1 consited of 20 rounds in which subjects observed the realizations of all of the observables and were asked to predict
the state. We had 5 observable variables called variable 1, 2, 3, 4 and 5. The variables were independent and Normally 
distributed with means and variances as described in TABLE.

>>>>

A finite number of observable random variables, $X =(x_1, x_2, ..., x_N)$, determine a binary state, $y$. 
The observable variables are independently distributed and each follows a Normal distribution with corresponding mean 
$\mu_i$ and standard deviation $\sigma_i$. The state is \emph{Red} (**R**) or \emph{Blue} (**B**) and is determined in 
the following way:

\begin{equation*}
y = \begin{cases}
\text{\textbf{B}} &\text{if $a_1x_1+...+a_Nx_N + K\geq0$}\\
\text{\textbf{R}} &\text{otherwise}
\end{cases}
\end{equation*}

@@

In each period, an agent observes the realization of all observable variables and predicts the state. Their flow payoff in 
period $t$ is 1 if they predict the state correctly, and it is zero otherwise. Denote the prediction by $\hat{y}$.

In order to make a prediction, the agent uses a map $M_t\to \{\textbf{B}, \textbf{R}\}$  where 
$M_t \subseteq \{x_1, x_2, ..., x_N\}$ is the set of variables that they choose to consider in period $t$. 
Our main object of study is the set of variables that the agent decides to consider in a given period, and the procedure by 
which the maping or the set are updated. We refer to the set of variables that are taken into consideration, $M_t$, as the agent's 
*mental model*, and assume that the mapping is linear.

A mental model is *simple* if it does not include all the relevant variables, 
*i.e.* if $M\subset \mathcal{R}$. And it is *complex* if it includes all relevant variables, $M=\mathcal{R}$. In 
addition, mental models can be ranked in terms of their simplicity by the number of relevant variables that they consider.

Whenever agents use simple models, they ignore some of the relevant information that is available to them.
In contrast, whenever agents use complex models, they use all the relevant information available. This means that when agents 
with complex mental models are restricted in terms of how many variables they can consider, their ability
to predict the state will be impaired. On the contrary, restricting the number of variables that an agent with a simple model can 
use, does not necessarily affect their ability to predict the state. In particular, if they are considering fewer variables than 
what the restriciton allows, their predictions should be as accurate as without the restriciton. These two effects are captured in 
predictions 1 and 1B respectively.

**Prediction 1:** If subjects use complex models, then allowing them to use an additional variable will always 
increase the frequency with which they make correct predictions.

**Prediction 1B:** If subjects use a simple mental model that considers $K$ variables, allowing them to use $K+1$ variables 
will not increase the frequency with which they make correct predictions.

### Optimal Rules
Conditional of using a particular mental model, the most accurate prediction rule that the agent can use is one that 
predicts the state to be **B** whenever, given the observables under consideration, **B** is more likely. 
And it predicts **R** whenever **R** is more likely. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the agent has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already. This provides an 
upper bound for how well any model can perform.

We refer to a prediction rule as an *Optimal Rule* if it is the rule that a fully rational agent would 
converge to with perfect Bayesian learning. In this context, Bayesian learning corresponds to using Bayes rule to 
learn the coefficients $a = (a_1, a_2, ..., a_N)$ as well as the constant $K$. This means that in an $N$ dimensional 
space, the agent must learn $N+1$ parameters which corresponds to learning the halfspace in which the state 
is **B**. The optimal rule for the complex model coincides with the true data-generating process when learning 
is perfect and the parameters are identified.

An optimal rule conditional on model $M$ is the rule that, when restricted to the 
variables being considered, does the best at predicting the state. In this case, the agent is learning only 
the parameters that pertain the variables in $M$ plus a constant term. The following proposition illustrates exactly 
what the optimal prediction rules look like when the agent is able to learn perfectly.

##### Proposition.
_Fix any model that considers $m\leq N$ variables and relabel the variable so that $x_1, ..., x_m$ are the variables 
that the model considers and $x_{m+1}, ..., x_N$ are the ones that the model does not consider. Relabel the 
coefficients $a = (a_1, ..., a_N)$ accordingly as well._

##### proof.
_Let $M:=a_1x_1+...+a_mx_m$ and $M^c = a_{m+1}x_{m+1}+...+a_Nx_N+k$ and define the latent variable $y^L := M +  M^c$. 
The optimal prediction rule for model $M$ predicts the state to be $\textbf{B}$ whenever 
$P[y^L|M]\geq \frac{1}{2}$ and predicts $\textbf{R}$ otherwise._

@@

With limited data, it is not guaranteed that the agent is able to perfectly learn the parameters required for the optimal rule. 
However, the notion of optimal rules is useful for determining a benchmark for how well each model can perform, as well as 
the scenarios in which we can expect polarization to arise. 
The next section uses the notion of optimal rules to formalize the concept of polarization and to establish the 
main prediction of the theory.

### Polarization
We take the definition of polarization from @Haghtalab2021: two agents are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one agent predicts the state to be **B** and the other 
predicts the state to be **R**.

Within out framework---as in @Haghtalab2021---polarization may persist even when agents have 
access to infinite data because they might be using different models of the world. 
Nevertheless using different models does not necessarily imply that beliefs will always be polarized, it must also be 
the case that the realization of observables falls in a polarization zone. By *polarization zone*, we  
mean that if the two models being used for prediction are $M_1$ and $M_2$, the realization of the 
observables $x=(x_1, ..., x_N)$ must be such that the optimal rule for $M_1$ makes a different prediction from 
that of $M_2$. The following 
2-dimensional example illustrates why the use of different models is necessary but not sufficient for polarization 
to arise.

##### Example
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2-1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As Figure \ref{example} makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. The regions in which the rules make different predictions
are the disagreement zones._

```{r example, out.width='50%', fig.align='center', fig.cap='\\label{example} Polarization regions for two single variable models'}
knitr::include_graphics(here("diagram_polariazation_2variables.png"))
```

For any pair of agents who observe the same realization of the observables and use different models, it is possible
to determine whether they will be polarized or not by looking at the predictions made by the optimal rule corresponding to their 
models. Notice that the 
polarizing observations---those that fall into a polarization zone---are different for each pair of models. Also notice that 
our definition of the polarization zone assumes that the agents are using optimal rules. This is a very strong assumption 
and will most likely not be the case that subjects in the experiment learn perfectly. This is a theoretical construct that allows
us to formalize Prediction 2.

@@

Subjects for the experimet were recruited from the CESS lab subject pool at NYU. 
We recruited `r data[, uniqueN(participant.code)]` undergraduate students who participated in this experiment in person 
durinf the summer and fall of 2023. The experiment was coded using o-tree @otree and it consisted of two parts:
Part 1 was meant to expose the subjects to the data generating process
and allow them to learn how to predict; part 2 was meant to elicit the models of the world that subjects might have
developed in part 1.

Part 1 consited of 20 rounds in which subjects observed the realizations of all of the observables and were asked to predict
the state. We had 5 observable variables called variable 1, 2, 3, 4 and 5. The variables were independent and Normally 
distributed with means and variances as described in TABLE.

### 2023-07-11 paper/paper.tex
edit framework writing

A finite number of observable random variables,
\(X =(x_1, x_2, ..., x_N)\), determine a binary state, \(y\). The
observable variables are independently distributed and each follows a
Normal distribution with corresponding mean \(\mu_i\) and standard
deviation \(\sigma_i\) with \(i \in \{1, 2, ..., N\}\).

The state is \emph{Red} (\textbf{R}) or \emph{Blue} (\textbf{B}) and is
determined in the following way: \begin{equation*}
y = \begin{cases}
\text{\textbf{B}} &\text{if $a_1x_1+...+a_Nx_N + K\geq0$}\\
\text{\textbf{R}} &\text{otherwise}
\end{cases}
\end{equation*}

@@

In each period, the decision-maker (DM) observes the realization of all
observable variables and predicts the state. Their flow payoff in period
\(t\) is 1 if they predict the state correctly, and it is zero
otherwise. Denote the prediction by \(\hat{y}\).

In order to make a prediction, the DM uses a map
\(f: M_t\to \{\textbf{B}, \textbf{R}\}\)\\
where \(M_t \subseteq \{x_1, x_2, ..., x_N\}\) is the set of variables
that they choose to consider in period \(t\). Our main object of study
is the set of variables that the DM decides to consider and not the
properties of the mapping from the variables to the state. We refer to
the set of variables that are taken into consideration, \(M_t\), as the
DM's \emph{model of the world}, and we abstract away from how the
variables are mapped into a prediction of the state.

A model of the world is said to be \emph{simple} if it does not include
all the relevant variables, \emph{i.e.} if \(M\subset \mathcal{R}\). And
it is complex if it includes all relevant variables, \(M=\mathcal{R}\).
Thus, whenever agents are using simple models, they are ignoring some of
the relevant variables even if the information is available to them. In
contrast, whenever agents are using complex models, they are using all
the relevant information that is available to them, so restricting the
number of variables that is available to them will impair their ability
to predict correctly. In the design of the experiment we will exploit
these differences in order to determine whether subjects are using
simple or complex models.

\textbf{Prediction 1:} If subjects are using simple model with \(K\)
variables, then allowing them to disclose \(K+1\) variables will not
increase the frequency with which they make correct predictions.

\textbf{Prediction 1B:} If subjects are using complex models, then
allowing them to disclose an additional variable will always increase
the frequency with which they make correct predictions.

@@

The most accurate prediction rule for this setting will be one that
predicts the state to be \textbf{B} whenever, given the observables,
\textbf{B} is more likely. And it predicts \textbf{R} whenever
\textbf{R} is more likely given the observables. Because we abstract
away from the procedure through which the rule is updated, we will make
predictions under the assumption that the DM has learned the parameters
perfectly. That is, they have seen enough information that, under an
updating procedure that converges to the truth---for example, correctly
specified regression model or Bayes rule---they would have converged to
the true parameters already.

We refer to a prediction rule as an \emph{Optimal Rule} if it is the
rule that a fully rational Decision maker would converge to with perfect
Bayesian learning. In this context, Bayesian learning corresponds to
using Bayes rule to learn the coefficients \(a = (a_1, a_2, ..., a_N)\)
as well as the constant \(K\). This means that in an \(N\) dimensional
space, the Decision Maker must learn \(N+1\) parameters which
corresponds to learning the halfspace in which the state is \textbf{B}.
The optimal rule for the complex model coincides with the true
data-generating process when learning is perfect.

Conditional on using a simple model, the optimal rule for that model is
the rule that, when restricted to the variables being considered, does
the best at predicting the state. In this case, the Decision Maker is
learning only the parameters that pertain the variables in the model.
Hence, for a model that considers \(K\) variables, the DM will have to
learn \(K+1\) parameters. The following proposition illustrates exactly
what the optimal prediction rules look like when the DM is able to
learn.

@@

\emph{Fix any model that considers \(k\leq N\) variables and relabel the
variable so that \(x_1, ..., x_K\) are the variables that the model
considers and \(x_{k+1}, ..., x_N\) are the ones that the model does not
consider. Relabel the coefficients \(a = (a_1, ..., a_N)\) accordingly
as well.}

@@

\emph{Let \(M:=a_1x_1+...+a_kx_k\) and
\(M^c = a_{k+1}x_{k+1}+...+a_Nx_N+k\) and define the latent variable
\(y^L := M + M^c\). The optimal prediction rule for a model that
considers the variables in \(M\) the state to be \(\textbf{B}\) whenever
\(P[y^L|M]\geq \frac{1}{2}\) and predicts \(\textbf{R}\) otherwise.}

@@

With limited data, it is not guaranteed that the DM is able to perfectly
learn the parameters required for the optimal rule. However, the notion
of optimal rules is useful for determining the scenarios in which we can
expect polarization to arise. The next section uses the notion of
optimal rules to formalize the concept of polarization and to establish
the main prediction of the theory.

@@

To study polarization, we take the definition of \citet{Haghtalab2021}:
two DMs are said to be polarized if, when observing the same realization
of observables, they predict the state to be different. That is, they
both observe the same draw of \(x=(x_1, ..., x_N)\) but one DM predicts
the state to be \textbf{B} and the other predicts the state to be
\textbf{R}.

Within out framework---as in \citet{Haghtalab2021}---polarization may
persist even when DMs have access to infinite data because they might be
using different models of the world in order to make their predictions.
Nevertheless using different models does not necessarily imply that
beliefs will always be polarized; it must also be the case that the
realization of observables falls in a \emph{disagreement zone}. By
disagreement zone, we simply mean that if the two models that are being
used for prediction are \(M_1\) and \(M_2\), the realization of the
observables \(x=(x_1, ..., x_N)\) must be such that \(M_1\) makes a
different prediction from \(M_2\). The following 2-dimensional example
illustrates why the use of different models is necessary but not
sufficient for polarization to arise.

@@

\emph{The optimal rule for the complex model coincides with the truth,
it predicts \textbf{B} if \(x_2+1\geq x_1\). In addition, there are two
simple models: one that considers only \(x_1\)---call it \(M_1\)---and
one that considers only \(x_2\)---call it \(M_2\). The optimal rules for
\(M_1\) and \(M_2\) are threshold rules that predict the state is
\textbf{B} if the value of \(x_i\) is below (above) the threshold
\(t_i := \mu_{-i} + k\) and the predicted state is \textbf{R} otherwise.
These prediction rules are illustrated in . As the figure makes clear,
both rules agree on the predicted state for some values of
\((x_1, x_2)\) and disagree for other values. Theregions in which the
rules make different predictions are the disagreement zones.}

For any pair of agents who observe the same realization of teh
observables and use different models, it is possible to determine
whether they will be polarized or not by looking at the predictions made
by their models. Notice that the polarizing observations---those that
fall into a polarization zone---are different for each pair of models.
Also notice that our definition of the polarization zone assumes that
the DMs are using optimal rules. This is a very strong assumption and
will most likely not be the case that subjects in the experiment learn
perfectly. However, the notion of optimal rules allows us to have a
clean definition of polarization and to make predictions about when
polarization will arise.

@@

Subjects for the experimet were recruited from the CESS lab subject pool
at NYU. We recruited 30 undergraduate students who participated in this
experiment in person at the CESS lab. The experiment was coded using
o-tree \citet{otree} and it consisted of two parts: Part 1 was meant to
expose the subjects to the data generating process and allow them to
learn how to predict; part 2 was meant to elicit the models of the world
that subjects might have developed in part 1.

Upon arrival subjects were randomly placed in pairs, however, they were
not informed of this fact. Part 1 consited of 20 rounds in which
subjects observed the realizations of all of the observables and were
asked to predict the state. We had 5 observable variables called
variable 1, 2, 3, 4 and 5. The variables were independent and Normally
distributed with means and variances as described in TABLE.

@@

In the first part of the experiment, subjects were asked to predict the
state given the realization of all the variables. Having access to all
the information they could predict the state perfectly if they managed
to learn the parameters of the hyperplane that determines the state.
Figure \ref{learningAll} shows the share of correct predictions across
rounds in part 1. We see that subjects are able to learn how to predict
the state at a rate that is higher than random. They initially predict
the state corectly about 50\% of the time, which is consistent with
random guessing. The share of correct guesses increases to 59\% by the
end of part 1. And this last number is significantly higher than 50\%
(p-value = \ensuremath{3\times 10^{-4}}). They continue to learn even
after the first 20 rounds when we look at the rounds in which subjects
had access to all the information. By the end of the experiment the rate
of correct predictions is 59\% which is still significantly higher than
50\% (p-value = 0)

>>>>

A finite number of observable random variables,
\(X =(x_1, x_2, ..., x_N)\), determine a binary state, \(y\). The
observable variables are independently distributed and each follows a
Normal distribution with corresponding mean \(\mu_i\) and standard
deviation \(\sigma_i\). The state is \emph{Red} (\textbf{R}) or
\emph{Blue} (\textbf{B}) and is determined in the following way:

\begin{equation*}
y = \begin{cases}
\text{\textbf{B}} &\text{if $a_1x_1+...+a_Nx_N + K\geq0$}\\
\text{\textbf{R}} &\text{otherwise}
\end{cases}
\end{equation*}

@@

In each period, an agent observes the realization of all observable
variables and predicts the state. Their flow payoff in period \(t\) is 1
if they predict the state correctly, and it is zero otherwise. Denote
the prediction by \(\hat{y}\).

In order to make a prediction, the agent uses a map
\(M_t\to \{\textbf{B}, \textbf{R}\}\) where
\(M_t \subseteq \{x_1, x_2, ..., x_N\}\) is the set of variables that
they choose to consider in period \(t\). Our main object of study is the
set of variables that the agent decides to consider in a given period,
and the procedure by which the maping or the set are updated. We refer
to the set of variables that are taken into consideration, \(M_t\), as
the agent's \emph{mental model}, and assume that the mapping is linear.

A mental model is \emph{simple} if it does not include all the relevant
variables, \emph{i.e.} if \(M\subset \mathcal{R}\). And it is
\emph{complex} if it includes all relevant variables, \(M=\mathcal{R}\).
In addition, mental models can be ranked in terms of their simplicity by
the number of relevant variables that they consider.

Whenever agents use simple models, they ignore some of the relevant
information that is available to them. In contrast, whenever agents use
complex models, they use all the relevant information available. This
means that when agents with complex mental models are restricted in
terms of how many variables they can consider, their ability to predict
the state will be impaired. On the contrary, restricting the number of
variables that an agent with a simple model can use, does not
necessarily affect their ability to predict the state. In particular, if
they are considering fewer variables than what the restriciton allows,
their predictions should be as accurate as without the restriciton.
These two effects are captured in predictions 1 and 1B respectively.

\textbf{Prediction 1:} If subjects use complex models, then allowing
them to use an additional variable will always increase the frequency
with which they make correct predictions.

\textbf{Prediction 1B:} If subjects use a simple mental model that
considers \(K\) variables, allowing them to use \(K+1\) variables will
not increase the frequency with which they make correct predictions.

@@

Conditional of using a particular mental model, the most accurate
prediction rule that the agent can use is one that predicts the state to
be \textbf{B} whenever, given the observables under consideration,
\textbf{B} is more likely. And it predicts \textbf{R} whenever
\textbf{R} is more likely. Because we abstract away from the procedure
through which the rule is updated, we will make predictions under the
assumption that the agent has learned the parameters perfectly. That is,
they have seen enough information that, under an updating procedure that
converges to the truth---for example, correctly specified regression
model or Bayes rule---they would have converged to the true parameters
already. This provides an upper bound for how well any model can
perform.

We refer to a prediction rule as an \emph{Optimal Rule} if it is the
rule that a fully rational agent would converge to with perfect Bayesian
learning. In this context, Bayesian learning corresponds to using Bayes
rule to learn the coefficients \(a = (a_1, a_2, ..., a_N)\) as well as
the constant \(K\). This means that in an \(N\) dimensional space, the
agent must learn \(N+1\) parameters which corresponds to learning the
halfspace in which the state is \textbf{B}. The optimal rule for the
complex model coincides with the true data-generating process when
learning is perfect and the parameters are identified.

An optimal rule conditional on model \(M\) is the rule that, when
restricted to the variables being considered, does the best at
predicting the state. In this case, the agent is learning only the
parameters that pertain the variables in \(M\) plus a constant term. The
following proposition illustrates exactly what the optimal prediction
rules look like when the agent is able to learn perfectly.

@@

\emph{Fix any model that considers \(m\leq N\) variables and relabel the
variable so that \(x_1, ..., x_m\) are the variables that the model
considers and \(x_{m+1}, ..., x_N\) are the ones that the model does not
consider. Relabel the coefficients \(a = (a_1, ..., a_N)\) accordingly
as well.}

@@

\emph{Let \(M:=a_1x_1+...+a_mx_m\) and
\(M^c = a_{m+1}x_{m+1}+...+a_Nx_N+k\) and define the latent variable
\(y^L := M + M^c\). The optimal prediction rule for model \(M\) predicts
the state to be \(\textbf{B}\) whenever \(P[y^L|M]\geq \frac{1}{2}\) and
predicts \(\textbf{R}\) otherwise.}

@@

With limited data, it is not guaranteed that the agent is able to
perfectly learn the parameters required for the optimal rule. However,
the notion of optimal rules is useful for determining a benchmark for
how well each model can perform, as well as the scenarios in which we
can expect polarization to arise. The next section uses the notion of
optimal rules to formalize the concept of polarization and to establish
the main prediction of the theory.

@@

We take the definition of polarization from \citet{Haghtalab2021}: two
agents are said to be polarized if, when observing the same realization
of observables, they predict the state to be different. That is, they
both observe the same draw of \(x=(x_1, ..., x_N)\) but one agent
predicts the state to be \textbf{B} and the other predicts the state to
be \textbf{R}.

Within out framework---as in \citet{Haghtalab2021}---polarization may
persist even when agents have access to infinite data because they might
be using different models of the world. Nevertheless using different
models does not necessarily imply that beliefs will always be polarized,
it must also be the case that the realization of observables falls in a
\emph{disagreement zone}. By disagreement zone, we simply mean that if
the two models that are being used for prediction are \(M_1\) and
\(M_2\), the realization of the observables \(x=(x_1, ..., x_N)\) must
be such that \(M_1\) makes a different prediction from \(M_2\). The
following 2-dimensional example illustrates why the use of different
models is necessary but not sufficient for polarization to arise.

@@

\emph{The optimal rule for the complex model coincides with the truth,
it predicts \textbf{B} if \(x_2+1\geq x_1\). In addition, there are two
simple models: one that considers only \(x_1\)---call it \(M_1\)---and
one that considers only \(x_2\)---call it \(M_2\). The optimal rules for
\(M_1\) and \(M_2\) are threshold rules that predict the state is
\textbf{B} if the value of \(x_i\) is below (above) the threshold
\(t_i := \mu_{-i} + k\) and the predicted state is \textbf{R} otherwise.
These prediction rules are illustrated in . As Figure \ref{example}
makes clear, both rules agree on the predicted state for some values of
\((x_1, x_2)\) and disagree for other values. The regions in which the
rules make different predictions are the disagreement zones.}

{\centering \includegraphics[width=0.5\linewidth]{../diagram_polariazation_2variables}

\caption{\label{example} Polarization regions for two single variable models}\label{fig:example}
\end{figure}

For any pair of agents who observe the same realization of the
observables and use different models, it is possible to determine
whether they will be polarized or not by looking at the predictions made
by the optimal rule corresponding to their models. Notice that the
polarizing observations---those that fall into a polarization zone---are
different for each pair of models. Also notice that our definition of
the polarization zone assumes that the agents are using optimal rules.
This is a very strong assumption and will most likely not be the case
that subjects in the experiment learn perfectly. This is a theoretical
construct that allows us to formalize Prediction 2.

@@

Subjects for the experimet were recruited from the CESS lab subject pool
at NYU. We recruited 30 undergraduate students who participated in this
experiment in person durinf the summer and fall of 2023. The experiment
was coded using o-tree \citet{otree} and it consisted of two parts: Part
1 was meant to expose the subjects to the data generating process and
allow them to learn how to predict; part 2 was meant to elicit the
models of the world that subjects might have developed in part 1.

Part 1 consited of 20 rounds in which subjects observed the realizations
of all of the observables and were asked to predict the state. We had 5
observable variables called variable 1, 2, 3, 4 and 5. The variables
were independent and Normally distributed with means and variances as
described in TABLE.

@@

In the first part of the experiment, subjects were asked to predict the
state given the realization of all the variables. Having access to all
the information they could predict the state perfectly if they managed
to learn the parameters of the hyperplane that determines the state.
Figure \ref{learningAll} shows the share of correct predictions across
rounds in part 1. We see that subjects are able to learn how to predict
the state at a rate that is higher than random. They initially predict
the state corectly about 50\% of the time, which is consistent with
random guessing. The share of correct guesses increases to 59\% by the
end of part 1. And this last number is significantly higher than 50\%
(p-value = \ensuremath{3\times 10^{-4}}). They continue to learn even
after the first 20 rounds when we look at the rounds in which subjects
had access to all the information. By the end of the experiment the rate
of correct predictions is 59\% which is still significantly higher than
50\% (p-value = 0).
